---
title: "An Introduction to species distribution models"
author: "Eduardo Arlé"
date: "22/12/2024"
output: html_document
---

## Load the packages we will need (if a package is still not in installed, use install.packages("PackageName") to install it).

```{r, print = FALSE}
library(rnaturalearth);library(sp);library(sf);library(plotfunctions)
library(rgbif);library(raster);library(bRacatus);library(geodata)
```


## Download species data from GBIF.

```{r}
koala <- occ_search(scientificName="Phascolarctos cinereus",
                      hasCoordinate=T)

koala <- koala[[3]]

koala_pts2 <- giveOcc(koala_pts, 
                         longitude = "decimalLongitude",
                         latitude = "decimalLatitude")

coordinates(koala_pts) <- ~decimalLongitude+decimalLatitude  #inform coordinates
  
#visualise
plotOcc(koala_pts2)
```


## Download bioclimatic variables.

```{r, eval = FALSE}
bioclim <- worldclim_global(var="bio", res=10, path = getwd())

class(bioclim)
nlayers(bioclim)

bioclim[[1]]

annual_mean_temp <- bioclim[[1]]

class(annual_mean_temp)

plot(annual_mean_temp)
```


## The geodata server is down for maintenance. We need to load the vars manually.

```{r}
setwd("/Users/carloseduardoaribeiro/Documents/CNA/Data/Variables/wc2-5")
vars <- lapply(list.files(pattern = '.bil$'), raster)
names(vars) <- gsub('.bil', '', list.files(pattern = '.bil$'))
```


## Crop the raster to the area of study

```{r}
annual_mean_temp <- vars$bio1

extent(annual_mean_temp)
extent(koala_pts)

study_area <- crop(annual_mean_temp, extent(koala_pts))

plot(study_area)
plot(koala_pts, add=T, pch=19, cex=.5)
```

## Visualising another layer of the raster stack, and crop the whole area of Australia

```{r}
annual_prec <- vars$bio12

plot(annual_prec)

australia_prec <- crop(annual_prec,extent(109,156,-44,-10))
australia_temp <- crop(annual_mean_temp,extent(109,156,-44,-10))

plot(australia_prec)
plot(koala_pts,add=T,pch=19,cex=.5)
```


## Extracting values from rasters at point locations

```{r}
australia_var <- stack(australia_prec, australia_temp)
australia_var

values <- extract(australia_var,koala_pts)
head(values)

class(values)
values <- as.data.frame(values)
class(values)

min_temp <- min(values$bio1)
max_temp <- max(values$bio1)

min_prec <- min(values$bio12)
max_prec <- max(values$bio12)
```


## Produce a very simple SDM for the Koala

```{r}
climatic_suitability <- australia_prec

for(i in 1:length(climatic_suitability[]))
{
  climatic_suitability[i] <- ifelse(australia_temp[i] > min_temp &
                            australia_temp[i] < max_temp &
                            australia_prec[i] > min_prec &
                            australia_prec[i] < max_prec,
                            1,0)
  
}

plot(climatic_suitability)
```
